---
title: "Data607 HW3 William Aiken"
author: "William Aiken"
date: "9/12/2021"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringr)
library(readr)
library(dplyr)
```

# Introduction
This homework assignment explores regex in R

### Problem 1
Using the 173 majors listed in fivethirtyeight.com’s College Majors dataset [link](https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/), provide code that identifies the majors that contain either "DATA" or "STATISTICS"

#### Solution

* Read in data from Github
* Create an new column to indicate if a major includes "DATA" or "STATISTICS" using 'str_detect'
* Filter on column to identify majors of interest

```{r}
file_path <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/majors-list.csv"

majors <- read_csv(file_path, show_col_types = FALSE)

majors <- majors %>% mutate(has_data_stats = as.numeric(str_detect(Major, "DATA|STATISTICS")))

majors %>% filter(has_data_stats == 1)
```

### Problem 2
Write code that transforms the data below:

[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"

[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  

[9] "elderberry"   "lime"         "lychee"       "mulberry"    

[13] "olive"        "salal berry"

Into a format like this:

c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")

#### Solution

* Put fruit in vector
* Use 'str_c' with ther parameter 'collapse' to collapse the vector into a single character string

```{r}
temp <- c("bell pepper", "bilberry","blackberry", "blood orange","blueberry", "cantaloupe", "chili pepper", "cloudberry","elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")
temp2 <- str_c(temp, collapse = ', ')
temp2
```

### Problem 3
Describe, in words, what these expressions will match:

    (.)\1\1
    "(.)(.)\\2\\1"
    (..)\1
    "(.).\\1.\\1"
    "(.)(.)(.).*\\3\\2\\1"

#### Solution
(.)\1\1
* First look for any character except a new line, repeat the whatever is in the first capture group.
* Finally, repeat what is in the first capture group.  
* This will find a character repeated 3 times, like "XXX"
```{r}
#I have escape the '\' to pass the regex as a string
str_detect("XXX", '(.)\\1\\1')
```

"(.)(.)\\2\\1"
* This expression is a string so the backslashes need to be escaped. 
* First look for any character except a new line.
* Second, look for for any character except a new line.  
* Then repeat what is in the second capture group. 
* Finally, repeat what is in the first capture group.
* This will find a 4 character palindrome, like "kook".
```{r}
str_detect("kook", "(.)(.)\\2\\1")
```

(..)\1
* First look for any 2 characters that aren't a new line.  
* Then repeat what is in the first capture group.
* This will find any two characters repeated, like "dodo"
```{r}
#got to escape '\'
str_detect("dodo", "(..)\\1")
```

"(.).\\1.\\1"
* This expression is a string so the backslashes need to be escaped. 
* First look first look for any character except a new line.
* Then look past any non-new line character and repeat the first capture group.  
* Then skip past any non-new line character and repeat the first capture group.  
* The last thing you do is evaluate the two characters we skipped over.
* This will find 3 repeated characters with an character separating them, like "babab"
```{r}
str_detect("babab", "(.).\\1.\\1")
```

"(.)(.)(.).*\\3\\2\\1"
* This expression is a string so the backslashes are escaped.  
* This will find any non-new line character and do that 2 more times.
* Then it skips over any number of non-new line characters and then repeats a search for the third capture group.  
* Then repeat the second capture group and finally repeat the first.
* This will find a string of characters where the first 3 and last 3 characters mirror each other like "wollow below"
```{r}
str_detect("wollow below", "(.)(.)(.).*\\3\\2\\1")
```

### Problem 4
Construct regular expressions to match words that:

    Start and end with the same character.
    Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)
    Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)

#### Solution 4
* Start and end with the same character.
* ^(.).*\1$
```{r}
#have to escape '\'
str_match('wow', "^(.).*\\1$")
```

Contain a repeated pair of letters (e.g. “church” contains “ch” repeated twice.)
* .*(..).*\1.*
```{r}
#have to escape '\'
str_match("church", ".*(..).*\\1.*")
```

Contain one letter repeated in at least three places (e.g. “eleven” contains three “e”s.)
* .*(.).*\1.*\1.*
```{r}
#have to escape '\'
str_match("eleven", ".*(.).*\\1.*\\1.*")
```

#Conclusion
Regex is a really powerful way to parse strings and can reduce time consuming data parse to a trivial task.

