---
title: "Data607_HW5_WilliamAiken.Rmd"
author: "William Aiken"
date: "9/26/2021"
output: html_document
---

# Introduction
### This is a creation and analysis of Israeli Vaccination data for Covid 19

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
library(readr)
library(stringr)
library(kableExtra)
library(dplyr)
```

# Methods
1. First we need to recreate the table of vaccination data
* create a data frame to populate with our vaccination data
* create column names based on the table provided
* create a row name called 'Age'
```{r}
covid_df <-
  data.frame(
    Age = as.character(),
    "Population.Not.Vaxed" = as.character(),
    "Population.Fully.Vaxed" = as.character(),
    "Severe.Cases.Not.Vax.per.100K" = as.character(),
    "Severe.Cases.Fully.Vax.per.100K" = as.character(),
    "Efficacy.vs.severe.disease" = as.character(), row.names = "Age"
  )
```

2. Populate the rows of our csv, with a primary row based on age stratification
* Create a secondary row with the age percentage
* Blank values are represented with 'NA'
```{r}
covid_df['<50',]<- c('1,116,834', '3,501,118', '43', '11', NA)
covid_df['<50%',]<- c('23.3%', '73.0%', NA, NA, NA)
covid_df['>50',]<- c('186,078', '2,133,516', '171', '290', NA)
covid_df['>50%',]<- c('7.9%%', '90.4%', NA, NA, NA)

covid_df %>% kbl %>% kable_classic
```

3. Save out our csv that we are going to commit to github
* We don't want to run this again
```{r}
#write.table(covid_df, "israeli_vaccination_data.csv", row.names=TRUE, sep = ",")
```

4. Load csv from github with readr
```{r}
file_path <- "https://raw.githubusercontent.com/catfoodlover/Data607/main/israeli_vaccination_data.csv"
covid_df2 <- read_csv(file_path, show_col_types = FALSE)
```

5.  readr saves the data as a tibble which doesn't support row names so we need to fix our data structrue
* convert to a data frame
* name the rows the first column's values
* save out a list of column names
* delete the first column
* split the last column which contains a comma separated list
* rename the columns with the saved list
* set all the strings 'NA' to NA
```{r}
covid_df2 <- as.data.frame(covid_df2)
rownames(covid_df2) <- covid_df2[, 1]

names_list <- names(covid_df2)
covid_df2[,1] <- NULL

covid_df2 <- covid_df2 %>% separate(Efficacy.vs.severe.disease, sep = ",", c("temp1", "temp2"))

colnames(covid_df2) <- names_list

covid_df2[covid_df2 == 'NA'] <- NA
```

6. We want to get the information in those secondary rows into columns
* use case_when to detect row names and put the age stratification into a new column called 'age_group'
* group by age_group and move those population %s into a new column and use the 'fill' function to populate all the rows
* remove that no longer needed secondary row
```{r}

covid_df2 <-
  covid_df2 %>% mutate(age_group = case_when(
    str_detect(row.names(.), "<50") ~ "<50",
    str_detect(row.names(.), ">50") ~ ">50"
  )) %>% group_by(age_group) %>% mutate(
    percent.not.vaxed = str_extract(Population.Not.Vaxed, ".*%"),
    percent.vaxed  = str_extract(Population.Fully.Vaxed, ".*%")
  ) %>% fill(percent.not.vaxed, .direction = c("up")) %>% fill(percent.vaxed, .direction = c("up")) %>% ungroup(.) %>% filter(!is.na(Severe.Cases.Not.Vax.per.100K))

covid_df2 %>% kbl() %>% kable_classic()
```

7. Now we can calculate our Efficacy vs severe disease
* we cast our severe cases per 100K to numeric and divide by 100K to get percentages
* we calculate Efficacy as 1 - (% Severe cases vaccinated per 100K)/(% Severe cases not vaccinated per 100K)
```{r}
covid_df2 <-
  covid_df2 %>% mutate(Efficacy.vs.severe.disease = 1 - ((as.numeric(Severe.Cases.Fully.Vax.per.100K)/100000)/(as.numeric(Severe.Cases.Not.Vax.per.100K)/100000)))
```

# Results

1.  Do we have enough information to calculate the total population?

### Yes we do, we know what percentages for both the under and over 50
* We divide either treatment arm by the percent population they represent to get the total population
* We sum the totals for both age groups to get the overall total population
```{r}
(under_50_pop <- 3501118/.73)

(over_50_pop <- 2113516/.904)

(total_pop <- under_50_pop + over_50_pop)
```

### This population represents people in Israel old enough to get vaccinated (12 years and older), total pop ~8.8 million

2. We have already calculated the efficacy
```{r}
temp <- covid_df2 %>% group_by(age_group) %>% summarise(Efficacy.vs.severe.disease)

kbl(temp) %>% kable_classic()
```

* We can see that vaccination has an efficacy of 74% in the under 50 group
* The vaccine has a surprising -69% efficacy in the over 50 group
* We will dig into explaining why this is in the next question

3. From your calculation are you able to compare the rates of severe disease in unvaccinated and vaccinated individuals?

```{r}

covid_df2 %>% select(age_group, Population.Not.Vaxed, percent.not.vaxed, Population.Fully.Vaxed, percent.vaxed, Efficacy.vs.severe.disease) %>% kbl() %>% kable_classic()

```

### My calculation doesn't allow me to measure vaccine efficacy for a couple of reasons. 
* Our percentages don't add up to 100% so I'm assuming partially vaccinated people and people vaccinated for less than 14 days are excluded?
* Those people should be in another category
* It also unclear to me what 'per 100k means', does it mean the entire cohort (>50/<50) or 100k within that age group and treatment arm?
* The fact that such a high percentage of older Israelis have been vaccinated it's not surprising that there are more breakthrough infections than infections in the unvaccinated



