---
title: "Data607 HW1"
author: "William Aiken"
date: "8/29/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(readr)  
library(knitr)
library(reshape2)
library(kableExtra)
library(gtsummary)
```

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
```

## Overview
### This is a further exploration of the halloween candy dataset taken from the *fivethirtyeight* article "The Ultimate Halloween Candy Power Ranking"
[link to The Ulitmate Halloween Candy Power Ranking](https://fivethirtyeight.com/videos/the-ultimate-halloween-candy-power-ranking/)

### In this article they explored the desirabiltiy of 85 common Halloween treats
#### They focused on properties of the treats to see if they could identify what made them desirable
* Chocolate
* Fruity
* Caramel
* Nuts
* Nougat
* Crispy
* Hard

```{r}
#Read in the data
file_path <- "https://raw.githubusercontent.com/catfoodlover/Data607/main/Data607_HW1_Data_WilliamAiken.csv"

candy_df <- read_csv(file_path, show_col_types = FALSE)

#Select columns of interest and fix names when necessary
candy_df <- candy_df %>% select(names = competitorname, choco = chocolate, fruity, caramel, nuts = peanutyalmondy, nougat, crispy = crispedricewafer, win_percent = winpercent)

#Character needs to be fixed
candy_df$names <- gsub( "Õ", "'", candy_df$names)

#Convert dataset from wide to long to make it easier to work with
temp <- melt(data = candy_df, id.vars = "names", measure.vars = c("choco", "fruity", "caramel", "nuts", "nougat", "crispy"), variable.name = "Property", value.name = "Status")

#Join win percentage back in
temp2 <- left_join(temp, candy_df %>% select(names, win_percent), by = "names")
```

### What is the mean win percentage of all properties?
* People like texture, nuts and crispy are stand outs both those are both properties that tend to go with chocolate the 3rd place property.
```{r}
temp2 %>% filter(Status == 1) %>% select(Property, win_percent) %>% tbl_summary(by = Property, 
  statistic = list(all_continuous() ~ "{mean} ({sd})"),
digits = all_continuous() ~ 1,
label = win_percent ~ "% Winner")

```


### what are the properties of the top bottom candies?
* Looks like people love chocolate.
```{r}
candy_df %>% arrange(desc(win_percent)) %>% slice_head(n = 10) %>% 
  count(choco, fruity, caramel, nuts, nougat, crispy) %>% kable() %>% kable_styling()
```

### What are the properties of the bottom ten treats?
* There a less clear message.  Fruity isn't popular but more importantly none of these are chocolate.
```{r}
candy_df %>% arrange(desc(win_percent)) %>% slice_tail(n = 10) %>% 
  count(choco, fruity, caramel, nuts, nougat, crispy) %>% kable() %>% kable_styling()
```

#Conclusion
* It looks like the sterotypes are true, people love chocolate.

##Future questions:
* Are people drawn to novelty?  Does how long a product been on the market predict a winning treat.
* Do certain combinatations of properties spell trouble or success?


