---
title: "DATA 606 Data Project Proposal"
author: "William Aiken"
output:
  pdf_document: default
---


### Data Preparation

```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
library(dplyr)
library(readr)
library(ggplot2)
library(kableExtra)
library(gtsummary)
# load data

Income <- readr::read_delim("https://raw.githubusercontent.com/catfoodlover/Data607/main/Income.csv", delim = ';', show_col_types = FALSE)
Obese <- readr::read_delim("https://raw.githubusercontent.com/catfoodlover/Data607/main/Obese.csv", delim = ';', show_col_types = FALSE)
Diagnosed <- readr::read_delim("https://raw.githubusercontent.com/catfoodlover/Data607/main/Diagnosed.csv", delim = ';', show_col_types = FALSE)

Diagnosed <- rename(Diagnosed, County = "County Name")
Diagnosed <- rename(Diagnosed, diabetic = "Percent/Rate")
Income <- rename(Income, Year = "Tax Year")

Income09 <- filter(Income, Year == '2009')
Income09 <- rename(Income09, income = "Average NY AGI of All Returns")

Obese <- rename(Obese, obesity = "Percentage/Rate")

#fix missing decimals
Diagnosed <- Diagnosed %>% mutate(diabetic = diabetic/10)


```


### Research question 

**You should phrase your research question in a way that matches up with the scope of inference your dataset allows for.**
Is there a correlation between obesity rates, average income and diabetic rates at the county level in New York State?


### Cases 

**What are the cases, and how many are there?**
These cases are the New York State reported county level obesity, income and diabetic rates.


### Data collection 

**Describe the method of data collection.**
The data is public health data that is collected by the state and income data collected by the state for tax purposes.



### Type of study 

**What type of study is this (observational/experiment)?**
This is an observational study.


### Data Source 

**If you collected the data, state self-collected. If not, provide a citation/link.**
This data comes from the NY.GOV site as part of their open data sets
[Obesity and Diabetes](https://health.data.ny.gov/Health/Community-Health-Obesity-and-Diabetes-Related-Indi/tchg-ruva)
[Income](https://data.ny.gov/Government-Finance/Average-Income-and-Tax-Liability-of-Full-Year-Resi/2w9v-ejxd)


### Dependent Variable

**What is the response variable? Is it quantitative or qualitative?**
The repsonse variable is diabetic rate which is quantitative.



### Independent Variable

**You should have two independent variables, one quantitative and one qualitative.**
My two independent variables are average income which is quantitative and obesity rate which is also quantitative.


### Relevant summary statistics 

**Provide summary statistics for each the variables. Also include appropriate visualizations related to your research question (e.g. scatter plot, boxplots, etc). This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r}


ggplot(data=Diagnosed, aes(diabetic)) + geom_histogram(bins = 50)


ggplot(data=Income09, aes(income)) + geom_histogram(bins = 50)


ggplot(data=Obese, aes(obesity)) + geom_histogram(bins = 50)


Diagnosed %>% select(diabetic) %>% tbl_summary(statistic = list(all_continuous() ~ "{mean} ({sd}), {min}, {max}"))

Income09 %>% select(income) %>% tbl_summary(statistic = list(all_continuous() ~ "{mean} ({sd}), {min}, {max}"))

Obese %>% select(obesity) %>% tbl_summary(statistic = list(all_continuous() ~ "{mean} ({sd}), {min}, {max}"))


```

