---
title: "Chess Tournament Data Wrangling"
author: "William Aiken"
date: "9/19/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(readr)  
library(knitr)
library(reshape2)
library(kableExtra)
library(gtsummary)
library(RPostgreSQL)
library(keyring)
library(readr)
library(tidyverse)
library(reshape2)
```

# Introduction
### We given a chess tournament data set as a text file and our goal is to convert this into a data structure containing the following elements:

* Player’s Name
* Player’s State
* Total Number of Points
* Player’s Pre-Rating
* Average Pre Chess Rating of Opponents

# Method
### First we import the text file which we have saved on Github and convert it to data frame
* We are using the 'readLines' funtion to import the text file, the parameter '-1' tells the function to read the entire file
* We then cast it as a data frame which gives us one column with all the data in it (We have some work to do).
* We use the 'separate' function from the 'tidyr' package to attemp to split the text into columns.  We are using the pipe as our delimiter.

```{r}
file_path <- "https://raw.githubusercontent.com/catfoodlover/Data607/main/tournamentinfo.txt"

chess_df <- readLines(file_path, n = -1)

chess_df <- as.data.frame(chess_df)

step <-
  tidyr::separate(
    data = chess_df,
    col = chess_df,
    into = c(
      "Pair_Num",
      "Player_Name",
      "Total_Pts",
      "Round_1",
      "Round_2",
      "Round_3",
      "Round_4",
      "Round_5",
      "Round_6",
      "Round_7"
    ),
    sep = "\\|",
  ) 

temp <- head(step)

kableExtra::kable(temp)
```

### We can see that this data set is going to require some work to get it to the final form of one event per row with aggregated opponent scores

* First we remove the rows without information with a simple filter statement
* We then split the data set into two data sets because each row contains two rows of information
* We create a flag called 'split' by casting 'Pair_Num' as numeric and filtering on whether it is missing or not

```{r}
step2 <- step %>% filter(!is.na(Player_Name))

step3 <- step2 %>% mutate(split = as.numeric(Pair_Num)) %>% filter(!is.na(split) == TRUE)

step4 <- step2 %>% mutate(split = as.numeric(Pair_Num)) %>% filter(is.na(split) == TRUE)

temp <- head(step3)
temp2 <- head(step4)
kableExtra::kable(temp)
kableExtra::kable(temp2)
```

### We now have two data frames, one with the player names, points, and record and a second data frame with player location and pre/post ranking
* Working the 2nd data frame, we split the column 'Player_Name into 'USCF_ID' and 'Pre_Post'
* We split "Pre_Post' again into to separate columns using '->' as our delimiter
* We use the 'gsub' function to clean up the rankings so they can later be cast as numeric
* We are also selecting off unused columns and rows
* Lastly we join this data frame back to the original data frame to get each players pre-tournament score associated with their 'Pair_Num'
```{r}
step5 <- step4 %>% separate(col = Player_Name, into = c("USCF_ID", "Pre_Post"), sep = "\\/") 

step6 <- step5 %>% separate(col = Pre_Post, into = c("Pre", "Post"), sep = "\\-\\>")

step7 <- step6 %>% mutate(Pre = gsub("R\\: ", "", Pre), Pre = gsub("P.*", "", Pre), Post = gsub("P.*", "", Post)) %>% select(Pair_Num, USCF_ID, Pre, Post)

step10 <- step7 %>% select(Location = Pair_Num, USCF_ID, Pre, Post) %>% filter(!is.na(Post), Post != "")

step11 <- bind_cols(step3, step10) %>% select(Pair_Num, Pre) %>% mutate(Pair_Num = str_trim(Pair_Num))

temp <- head(step10)
kableExtra::kable(temp)
```

### We turn our attention back to the first data set and convert it from wide to long for ease of calculations

* We use the 'melt' function to create one row per tournament round per player
* We use 'seperate' to split the the round results from the opponent 'Pair Number'
* We left join our opponent's pre-tournament ranking we created in the last code block
* Now we can filter on meaningful round results (Win, Lose, Draw) and group by player and find the average ranking of their opponents
* It's important to note that this gives us meaningful results for both players who played every round and players who had non-game rounds
* We create a temporary table to join player location into our final dataset by 'Pair Num'
* We trim unnecessary spaces from character strings to allow for more accurate joining

```{r}

step8 <- melt(data = step3, id.vars = "Player_Name", measure.vars = c("Round_1","Round_2","Round_3","Round_4","Round_5","Round_6","Round_7"))

step9 <- step8 %>% separate(col = value, into = c("Result", "Opponent"), sep = "  ") %>% mutate(Opponent = str_trim(Opponent), Result = str_trim(Result))

step12 <- left_join(step9, step11, by = c("Opponent" = "Pair_Num"))

step13 <-
  step12 %>% filter(Result %in% c("W", "L", "D")) %>% group_by(Player_Name) %>% mutate(avg_opp = round(sum(as.numeric(Pre)) /
                                                                                                         n())) %>% ungroup()

step14 <- bind_cols(step3, step10) %>% select(Pair_Num, Location) %>% mutate(Pair_Num = str_trim(Pair_Num))


step15 <- step3 %>% mutate(Pair_Num = str_trim(Pair_Num,), Player_Name = str_trim(Player_Name)) %>% select(Pair_Num, Player_Name, Total_Pts)
```

### Now we can put all the pieces together to create a final data frame to be uploaded into a database

*  We take our calculated opponent averages and split them off into a temporary table, with 'Player_Name' as the key
* We join all the required fields together, using 'Pair_Num' or 'Player_Name' as the key
```{r}

step16 <- step13 %>% mutate(Player_Name = str_trim(Player_Name)) %>% select(Player_Name, avg_opp) %>% distinct()

final <- left_join(step15, step14, by = c("Pair_Num"))

final <- left_join(final, step11, by = c("Pair_Num"))

final <- left_join(final, step16, by = c("Player_Name"))

final <- final %>% select(Player_Name, Player_State = Location, Total_Pts, Pre_Rating = Pre, Opponent_Avg_Pre = avg_opp)

temp <- head(final)

kableExtra::kable(temp)

```

### Now we can upload the final data frame to bit.io a free server

* Create empty table
* Use INSERT statement in postgres SQL to load data
* Create SELECT statement in postgres SQL to get table

```{r}

#get my password for bit.i
password <- key_get("bit.io", "catfoodlover_demo_db_connection")

#connect to bit.io
con <- dbConnect(RPostgreSQL::PostgreSQL(), dbname = 'bitdotio', 
                 host = 'db.bit.io',
                 port = 5432,
                 user = 'bitdotio',
                 password = password)


#Create movies table
DBI::dbSendQuery(con, 'CREATE TABLE IF NOT EXISTS "catfoodlover/demo_repo"."chess" (
  player_name TEXT,
  player_state TEXT,
  total_points DOUBLE PRECISION,
  pre_rating BIGINT,
  opponent_rating_avg BIGINT
)')

#Insert data into table
DBI::dbSendQuery(con, 'INSERT INTO "catfoodlover/demo_repo"."chess"  VALUES
(\'GARY HUA\',   \'ON\' ,6.0  , 1794   ,1605),
(\'DAKSHESH DARURI\',   \'MI\' ,6.0  , 1553   ,1469),
(\'ADITYA BAJAJ\',   \'MI\' ,6.0  , 1384   ,1564),
(\'PATRICK H SCHILLING\',   \'MI\' ,5.5  , 1716   ,1574),
(\'HANSHI ZUO\',   \'MI\' ,5.5  , 1655   ,1501),
(\'HANSEN SONG\',   \'OH\' ,5.0  , 1686   ,1519),
(\'GARY DEE SWATHELL\',   \'MI\' ,5.0  , 1649   ,1372),
(\'EZEKIEL HOUGHTON\',   \'MI\' ,5.0  , 1641,1468),
(\'STEFANO LEE\',   \'ON\' ,5.0  , 1411   ,1523),
(\'ANVIT RAO\',   \'MI\' ,5.0  , 1365   ,1554),
(\'CAMERON WILLIAM MC LEMAN\',   \'MI\' ,4.5  , 1712   ,1468),
(\'KENNETH J TACK\',   \'MI\' ,4.5  , 1663   ,1506),
(\'TORRANCE HENRY JR\',   \'MI\' ,4.5  , 1666   ,1498),
(\'BRADLEY SHAW\',   \'MI\' ,4.5  , 1610   ,1515),
(\'ZACHARY JAMES HOUGHTON\',   \'MI\' ,4.5  , 1220,1484),
(\'MIKE NIKITIN\',   \'MI\' ,4.0  , 1604   ,1386),
(\'RONALD GRZEGORCZYK\',   \'MI\' ,4.0  , 1629   ,1499),
(\'DAVID SUNDEEN\',   \'MI\' ,4.0  , 1600   ,1480),
(\'DIPANKAR ROY\',   \'MI\' ,4.0  , 1564   ,1426),
(\'JASON ZHENG\',   \'MI\' ,4.0  , 1595   ,1411),
(\'DINH DANG BUI\',   \'ON\' ,4.0  , 1563,1470),
(\'EUGENE L MCCLURE\',   \'MI\' ,4.0  , 1555   ,1300),
(\'ALAN BUI\',   \'ON\' ,4.0  , 1363   ,1214),
(\'MICHAEL R ALDRICH\',   \'MI\' ,4.0  , 1229   ,1357),
(\'LOREN SCHWIEBERT\',   \'MI\' ,3.5  , 1745   ,1363),
(\'MAX ZHU\',   \'ON\' ,3.5  , 1579   ,1507),
(\'GAURAV GIDWANI\',   \'MI\' ,3.5  , 1552   ,1222),
(\'SOFIA ADINA STANESCU-BELLU\',   \'MI\' ,3.5  , 1507   ,1522),
(\'CHIEDOZIE OKORIE\',   \'MI\' ,3.5  , 1602,1314),
(\'GEORGE AVERY JONES\',   \'ON\' ,3.5  , 1522   ,1144),
(\'RISHI SHETTY\',   \'MI\' ,3.5  , 1494   ,1260),
(\'JOSHUA PHILIP MATHEWS\',   \'ON\' ,3.5  , 1441   ,1379),
(\'JADE GE\',   \'MI\' ,3.5  , 1449   ,1277),
(\'MICHAEL JEFFERY THOMAS\',  \'MI\' ,3.5  , 1399   ,1375),
(\'JOSHUA DAVID LEE\',   \'MI\' ,3.5  , 1438   ,1150),
(\'SIDDHARTH JHA\',   \'MI\' ,3.5  , 1355   ,1388),
(\'AMIYATOSH PWNANANDAM\',   \'MI\' ,3.5  ,  980,1385),
(\'BRIAN LIU\',   \'MI\' ,3.0  , 1423   ,1539),
(\'JOEL R HENDON\',   \'MI\' ,3.0  , 1436,1430),
(\'FOREST ZHANG\',   \'MI\' ,3.0  , 1348   ,1391),
(\'KYLE WILLIAM MURPHY\',  \'MI\' ,3.0  , 1403,1248),
(\'JARED GE\',   \'MI\' ,3.0  , 1332   ,1150),
(\'ROBERT GLEN VASEY\',   \'MI\' ,3.0  , 1283   ,1107),
(\'JUSTIN D SCHILLING\',   \'MI\' ,3.0  , 1199   ,1327),
(\'DEREK YAN\',   \'MI\' ,3.0  , 1242   ,1152),
(\'JACOB ALEXANDER LAVALLEY\',   \'MI\' ,3.0  ,  377,1358),
(\'ERIC WRIGHT\',   \'MI\',2.5  , 1362   ,1392),
(\'DANIEL KHAIN\',   \'MI\' ,2.5  , 1382   ,1356),
(\'MICHAEL J MARTIN\',   \'MI\' ,2.5  , 1291,1286),
(\'SHIVAM JHA\',   \'MI\' ,2.5  , 1056   ,1296),
(\'TEJAS AYYAGARI\',   \'MI\' ,2.5  , 1011   ,1356),
(\'ETHAN GUO\',   \'MI\' ,2.5  ,  935   ,1495),
(\'JOSE C YBARRA\',   \'MI\' ,2.0  , 1393   ,1345),
(\'LARRY HODGE\',   \'MI\' ,2.0  , 1270   ,1206),
(\'ALEX KONG\',   \'MI\' ,2.0  , 1186   ,1406),
(\'MARISA RICCI\',   \'MI\' ,2.0  , 1153   ,1414),
(\'MICHAEL LU\',   \'MI\',2.0  , 1092   ,1363),
(\'VIRAJ MOHILE\',   \'MI\' ,2.0  ,  917   ,1391),
(\'SEAN M MC CORMICK\',   \'MI\' ,2.0  ,  853   ,1319),
(\'JULIA SHEN\',   \'MI\' ,1.5  ,  967   ,1330),
(\'JEZZEL FARKAS\',   \'ON\' ,1.5  ,  955,1327),
(\'ASHWIN BALAJI\',   \'MI\' ,1.0  , 1530   ,1186),
(\'THOMAS JOSEPH HOSMER\',   \'MI\' ,1.0  , 1175   ,1350),
(\'BEN LI\',   \'MI\' ,1.0  , 1163   ,1263);')

```

# Results
```{r}

d <- dbGetQuery(con, 'SELECT * FROM "catfoodlover/demo_repo"."chess";')


kableExtra::kable(d, col.names = c("Player’s Name", "Player’s State", "Total Number of Points", "Player’s Pre-Rating",  "Average Pre Chess Rating of Opponents"))
dbGetQuery(con, 'DROP TABLE "catfoodlover/demo_repo"."chess";')

```

# Conclusion

### Getting this data into its final form of one row per player with aggregated opponent pre-tournament ranking required a number of different data wrangling skills

* subsetting to split the data set into easier to manipulate pieces
* regex to split fields and remove unnecessary characters 
* reshaping to convert data from wide to long
* group by to all for aggregation of stratified data
* SQL to load data into data base for ease of access
* This made me appreciate data already structured one event per row




